Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-11-20 paper/main.tex
changes to impact geometry

    To understand how the impacted particle is perturbed, one needs to compute its change in momentum, which is determined by integrating the force acting on the particle over the duration of the interaction. A useful approximation for this change in momentum \paola{(per unit mass)} is the force at the closest approach multiplied by an estimate of the interaction time:
    \begin{equation} \label{eq:change_in_momentum} \Delta p \approx \text{Force} \times \text{interaction time} = \frac{GMm}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation}where $M$ is the mass of the perturber, $m$ is the mass of the particle, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. \paola{ATTENTION : in the last formula there is no dependence on $m$, as  for a change in momentum per unit mass (hence my comment few lines before). If you want to keep it as momentum per unit mass, that is a change in velocity, you should also remove $m$ from the other expression (i.e. GMm)}

@@

    Visual inspection of the five key impact parameters ($M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$) did not reveal a clear distinction between flybys that create gaps and those that do not. Due to the clutter, we omit this graph and instead present... In Fig.\ref{fig:impact_geometry_statistics} we report the result of this analysis, by showing the distribution of these impact variables for fly-bys which give rise to gaps and compare it to the distributions of fly-bys which do not lead to gap formation. Note that in this figure, we show the total relative velocity rather than separating parallel and perpendicular components, because from a visual inspection, we could not find any specific trends when the two velocity components were plotted separately. We also exclude the characteristic cluster radius, as it showed little correlation with the results, likely due to the narrow range of globular cluster radii (see Fig.\ref{fig:mass_size_plane}). This factor might be more significant for dark matter subhalos, where size variation is greater.

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone (from Eq.~\ref{eq:change_in_momentum}) cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 200~pc. The stream widths are roughly 200~pc, with a radial thickness\paola{what do you mean by radial thickness?} of about 100~pc. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim7 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, massive globulars with masses greater than $10^6 \rm{M}_\odot$ are not effective in producing gaps if they pass close to Pal~5 stream with an impact parameter $b > 200$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps. \paola{Attention: there is something I do not understand in Fig 5: the range of $\delta v$ values seems not the same in the first row and in the last row (left panel). Is there anything to clarify/correct?}

@@

  Once all the key impact parameters estimated, we can use Eq~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Pal~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities (there is a factor of about 100 between the smallest and largest changes), with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$. \paola{I suggest to use $\Delta p$ instead of $\Delta v$ in the xlabel of Fig.~\ref{fig:deltap}, to avoid confusion with $\delta v$ of Fig.\ref{fig:impact_geometry_statistics}.}\\

  In addition to characterizing the parameters governing cluster encounters with the stream, since we know which clusters have produced gaps on the tail of Pal5, we can also verify their orbital and structural properties.  This is what is shown in Fig~\ref{fig:mass_size_plane}, where we first show the mass and size (i.e. half-mass radius) of the clusters that cause gaps on the tails of Pal5, dividing them from those that do not. As can be seen, no cluster with mass below $10^5 \rm{M}_\odot$ causes gaps on Pal~5 stream, and all perturbers, except xx \paola{add which cluster has a $r_h > 10$pc} have a  half-mass radius below 10~pc. Even more interesting is their distribution in the E-Lz plane, which shows that the clusters that cause gaps are on both direct and retrograde orbits (respectively, negative and positive values of $L_z$), but all confined to an energy interval between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is because only clusters within Palomar 5's orbital space can interact with the stream: clusters with higher orbital energies tend to have larger apocenter than that of Pal~5 and thus spend most of their time away from Palomar 5's orbital volume. Finally,  it is worth commenting on the location of impacts with the stream, whether they occur when Pal~5 is close to its pericenter or not. As it can be seen from Fig.xx (bottom panel) encounters can occur at all orbital phases of Pal~5, when it is close to pericenter, but also very far away from it, at the outskirts of its orbital space. However, when taken all together, the location of gap creating impacts shows a strong negative correlation  with the galactocentric radius $r$, with the number $N$ of encounters favorable to gap formation  going as $N = -2.5r + 50$ (with a Pearson coefficient of -0.86). While there are more clusters near the galactic center, orbits naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster galactic number density outweighs these factors in determining the number of gaps.

>>>>

    To understand how the impacted particle is perturbed, one needs to compute its change in momentum, which is determined by integrating the force acting on the particle over the duration of the interaction. A useful approximation for this change in momentum \paola{(per unit mass)} is the force at the closest approach multiplied by an estimate of the interaction time:
    \begin{equation} \label{eq:change_in_momentum} \Delta p \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{\delta v} = 2\frac{GM}{b \delta v}, \end{equation}where $M$ is the mass of the perturber, $b$ is the impact parameter, $\delta v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant.

@@

    Visual inspection of the five key impact parameters ($M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$) did not reveal a clear distinction between flybys that create gaps and those that do not. Therefore, we only present the the quantities from Eq.~\ref{eq:change_in_momentum} in Fig.~\ref{fig:impact_geometry_statistics}. 
    Note that in this figure, we show the total relative velocity rather than separating parallel and perpendicular components, because from a visual inspection, we could not find any specific trends when the two velocity components were plotted separately. We also exclude the characteristic cluster radius, as it showed little correlation with the results, likely due to the narrow range of globular cluster radii (see Fig.\ref{fig:mass_size_plane}). This factor might be more significant for dark matter subhalos, where size variation is greater.

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 200~pc. The stream widths are roughly 200~pc. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim7 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, massive globulars with masses greater than $10^6 \rm{M}_\odot$ are not effective in producing gaps if they pass close to Pal~5 stream with an impact parameter $b > 200$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps.

@@

    Once all the key impact parameters estimated, we can use Eq~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Pal~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities (there is a factor of about 100 between the smallest and largest changes), with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$.

    In addition to characterizing the parameters governing cluster encounters with the stream, since we know which clusters have produced gaps on the tail of Palomar~5, we can also verify their orbital and structural properties. This is what is shown in Fig~\ref{fig:mass_size_plane}, where we first show the mass and size (i.e. half-mass radius) of the clusters that cause gaps on the tails of Pal5, dividing them from those that do not. As can be seen, no cluster with mass below $10^5 \rm{M}_\odot$ causes gaps on Pal~5 stream, and all perturbers, except xx \paola{add which cluster has a $r_h > 10$~pc} have a half-mass radius below 10~pc. Even more interesting is their distribution in the E-Lz plane, which shows that the clusters that cause gaps are on both direct and retrograde orbits (respectively, negative and positive values of $L_z$), but all confined to an energy interval between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is because only clusters within Palomar 5's orbital space can interact with the stream: clusters with higher orbital energies tend to have larger apocenter than that of Pal~5 and thus spend most of their time away from Palomar 5's orbital volume. Finally, it is worth commenting on the location of impacts with the stream, whether they occur when Pal~5 is close to its pericenter or not. As it can be seen from Fig.\ref{fig:mass_size_plane} (bottom panel) encounters can occur at all orbital phases of Pal~5, when it is close to pericenter, but also very far away from it, at the outskirts of its orbital space. However, when taken all together, the location of gap creating impacts shows a strong negative correlation  with the galactocentric radius $r$, with the number $N$ of encounters favorable to gap formation  going as $N = -2.5r + 50$ (with a Pearson coefficient of -0.86). While there are more clusters near the galactic center, orbits naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster galactic number density outweighs these factors in determining the number of gaps.
